---
title: "Spotify"
author: "Tracy Mei"
date: ""
output:
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, message=FALSE}
# packages used in our report
library(tidyverse)
library(readr)
library(ggplot2)
library(dplyr)
```


```{r import, message=FALSE}
# importing the Spotify dataset
spotify <- read_csv("Spotify-2000.csv")
```
# Introduction

## Data

```{r viewdata}
# Inspecting our dataset
head(spotify, 10)
```

Add desciription of data including variable information

## Exploratory graphs

```{r graph1}
spotify <- spotify |> select(Index, Title, Artist, Energy, Danceability, `Length (Duration)`, Popularity)
#i will create some visualizations here..
ggplot(spotify, aes(x = Energy, y = Danceability))+
  geom_col()
ggplot(spotify, aes(x = Energy, y = Popularity)) +
  geom_point(alpha = 0.6) +
  geom_smooth (method= 'lm', se = FALSE, color = "blue") +
  labs(title = "Does Higher Energy Lead to More Popular Songs?",
       x = "Energy",
       y = "Popularity") +
  theme_minimal()
#this one is comparing Energy and Popularity. so as we can see, popularity increases as energy level increase.
```



```{r graph2}
ggplot(spotify, aes(x = Danceability, y = Popularity)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  labs(title = "Does Danceability Affect Song Popularity?",
       x = "Danceability", y = "Popularity") +
  theme_minimal()
#This one compares danceability and popularity. it also show us same relationship with energy
```



```{r graph3}
ggplot(spotify, aes(x = Energy, y = Danceability)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  labs(title = "Relationship Between Energy and Danceability",
       x = "Energy", y = "Danceability") +
  theme_minimal()
#following code explains the relationship between danceability and energy. there is also a positive relation between them which can be further analysed.
```



```{r graph4}
spotify$Length <- as.numeric(spotify$`Length (Duration)`)

spotify %>%
  filter(`Length (Duration)` < 1000) %>%
  ggplot(aes(x = `Length (Duration)`, y = Popularity)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "purple") +
    labs(title = "Does Song Length Impact Popularity?",
         x = "Song Length (sec)", y = "Popularity") +
    theme_minimal()
#this one also shows that as song length decrease, popularity increase but we need some tests i think that may not be so valid.
```



```{r graph5}
ggplot(spotify) +
  geom_smooth(aes(x = Popularity, y = Energy,       color = "Energy"),       method = "lm", se = FALSE) +
  geom_smooth(aes(x = Popularity, y = Danceability, color = "Danceability"), method = "lm", se = FALSE) +
  geom_smooth(aes(x = Popularity, y = `Length (Duration)`,      color = "Length"),       method = "lm", se = FALSE) +
  scale_color_manual("Feature",
                     values = c("Energy"="goldenrod",
                                "Danceability"="tomato",
                                "Length"="steelblue")) +
  labs(title = "Feature vs Popularity Trends",
       x     = "Popularity",
       y     = "Feature value") +
  theme_minimal()
#everything in one graph. hope it helps. it took like 1 hour to me lol..
```

Add distribution graphs and descriptions.

## Research question

Choose the dependent variable and define research question.


# Linear Model

## Data Split

Split the data into training, validation and test sets.

## Model selection

Use methods like Lasso and best subset selection to choose the best regressors

## Results

Include the chosen regression summary, test MSE, and interpret results.

# Conclusion

Conclude, state limitations, etc.

# Contributions



AI statement:

